---
title: "SISMID Spatial Statistics in Epidemiology and Public Health 
\\

2016 R Notes: Infectious Disease Data"
author: |
  | Jon Wakefield
  | Departments of Statistics, University of Washington
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    keep_tex: yes
  slidy_presentation: default
---
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(collapse=TRUE, fig.align='center', tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=40), warning=FALSE,message=FALSE, cache=TRUE)
```

## Lower Saxony Measles Data

As a first example of the Held et al. (2005) approach we examine data on measles considered in this paper (actually an updated version).

These data are in the ``surveillance`` package.

The data consist of weekly measles counts over 2001 and 2002, for each of 17 administrative district  Weser-Ems region of Lower Saxony, Germany.


Included in the dataset are a $17 \times 17$ matrix of 0/1 entries indicating which areas share a common boundary.

There is also the population that is contained in each area, and various other data including vaccination information.

## Lower Saxony Measles: Reading in the Data

``data("measlesWeserEms")`` is of ``sts`` class.

The data object also contains a map of the region, as a ``SpatialPolygonsDataFrame``
\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50)}
library(surveillance)
data("measlesWeserEms")
```


## Map of total counts

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.height=3.0,fig.width=3.0}
plot(measlesWeserEms, type=observed~unit)
```

## Map of total counts with labels

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.height=3,fig.width=3}
plot(measlesWeserEms, type=observed~unit,
     population=measlesWeserEms@map$POPULATION / 100000,
     labels=list(labels="GEN", cex=0.3, font=3))
```

## Time series of all data

\small
```{r,collapse=TRUE, fig.height=3,fig.width=4.5,echo=TRUE}
plot(measlesWeserEms,type=observed~time,legend.opts=NULL)
```

## Time series with areas of all zeroes excluded

Two areas contain all zeroes

\small
```{r,collapse=TRUE, fig.show="hide",echo=TRUE}
measlesWeserEms15 <- measlesWeserEms[,colSums(observed(measlesWeserEms)) > 0]
plot(measlesWeserEms15)
```

## 

\small
```{r,collapse=TRUE, fig.height=4.0,fig.width=4.5,echo=FALSE}
plot(measlesWeserEms15)
```

## Animation

The commands 

``library(gridExtra)``

``animate(measlesWeserEms)``

produce an animation.

Try them yourself!

## Model framework

We have a negative binomial model with
$$E[Y_{it} | \mu_{it}] = \mu_{it}$$
and
$$var(Y_{it} | \mu_{it}] = \mu_{it}(1+\psi \mu_{it})$$
so that $\psi=0$ corresponds to a Poisson model.

## First model

We first fit the model
$$\mu_{it} = \underbrace{\lambda^{\mbox{\tiny{AR}}} }_{\exp(\alpha_0^{\mbox{\tiny{AR}}})} y_{i, t-1}  + N_{i} \lambda^{\mbox{\tiny{EN}}}_{t},$$
with endemic term:
$$
\log(\lambda^{\mbox{\tiny{EN}}}_{t} ) = \alpha_0^{\mbox{\tiny{EN}}}+\alpha_1 t + \gamma \sin(\omega t) + \delta \cos(\omega t)$$
where 

* $\lambda^{\mbox{\tiny{AR}}}$ is the epidemic force,

*  $N_{t}$ are  population counts in area $i$,

* $\lambda^{\mbox{\tiny{EN}}}_{t}$ is the endemic term,

* $\alpha_1$ is a slope parameter describing the large scale endemic temporal trend,

* $\gamma$  and $\delta$ are seasonal parameters and  do not vary across areas, $\omega = (2\pi )/52$.


## First model

``NegBin1`` gives a single overdisperion parameter, i.e., common to all areas.

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50)}
f.end1 <- addSeason2formula(f = ~ 1 +  t, S=1, period=52)
measles.mod1 <- list(
  ar = list(f = ~ 1),  
  end = list(f = f.end1, offset = population(measlesWeserEms)),  
  family = "NegBin1")  
results.mod1 <- hhh4(measlesWeserEms,control=measles.mod1)
confint(results.mod1,parm="overdisp")
```    

This interval suggests a Poisson model, with $\psi=0$ would 

## Results for model 1

\scriptsize
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
summary(results.mod1)
```           
   
   
## Results for model 1

\small
``idx2Exp = 1:4`` gives exponentiated parameters and ``amplitudeShift`` tramsforms to the amplitude and phase shift.

\scriptsize
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"} 
summary(results.mod1, idx2Exp = 1:4, amplitudeShift = TRUE)
```

## Plot of endemic component

\small
```{r,collapse=TRUE, fig.height=2.5,fig.width=3}
# Plot for the 7th area:
plot(results.mod1,type="season",components="end",main="")
```

## A second model

We now add a neigborhood term, with simple binary weights given by
$$\mu_{it} = \underbrace{\lambda^{\mbox{\tiny{AR}}} }_{\exp(\alpha_0^{\mbox{\tiny{AR}}})} y_{i, t-1} + \underbrace{\lambda^{\mbox{\tiny{NE}}}}_{\exp(\alpha_0^{\mbox{\tiny{NE}}})}\sum_{j \in \mbox{\small{ne}}(i)} y_{j,t-1} + N_{it} \lambda^{\mbox{\tiny{EN}}}_{it},$$
with endemic term:
$$
\log(\lambda^{\mbox{\tiny{EN}}}_{it} ) = \alpha_0^{\mbox{\tiny{EN}}}+\alpha_1 t + \gamma \sin(\omega t) + \delta \cos(\omega t)$$
where 

* $\lambda^{\mbox{\tiny{AR}}}$ is the epidemic force,

* $\lambda^{\mbox{\tiny{NE}}}$ is the neighborhood effect,

*  $N_{it}$ are (possibly standardized) population counts in area $i$ at time $t$,

* $\lambda^{\mbox{\tiny{EN}}}_{it}$ is the endemic term,

* $\gamma$  and $\delta$ are seasonal parameters and  do not vary across areas, $\omega = (2\pi )/52$.


## Lower Saxony Measles Data

The neighbourhood matrix is  a distance matrix (neighbourhood orders)

``wji`` is a standardized weight 

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50)}
wji <- neighbourhood(measlesWeserEms)/rowSums(neighbourhood(measlesWeserEms))
measles.mod2 <- list(ar = list(f = ~ 1), 
    ne = list(f = ~ 1, weights = wji),  
    end = list(f = f.end1, offset = population(measlesWeserEms)),  
    family = "NegBin1")  
results.mod2 <- hhh4(measlesWeserEms,control=measles.mod2)
```    

## Lower Saxony Measles Data

\scriptsize
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
summary(results.mod2)
```           
                                                                      
## Lower Saxony Measles Data

Hence, the mean model is
\begin{small}
\begin{eqnarray*}
E[Y_{i,t+1}|y_{it},y_{jt}] &=& \exp( -0.38) y_{it} + \exp(-25.05) \sum_{j \in \mbox{\small{ne}}(i)} y_{jt} \\&+&
\exp \left[ 0.40 + 0.0029 \times t + 1.0 \sin(\omega t) -0.53  \cos(\omega t) \right]
\end{eqnarray*}
\end{small}
The variance is
\begin{small}
$$var (Y_{i,t+1}|y_{it} ) = E[Y_{it}|y_{it}] (1+ 2.23 \times E[Y_{it}|y_{it}] ).$$
\end{small}

The fits are different in each of the areas due to the two epidemic components (AR and NE).

## Lower Saxony Measles Data

\small
```{r,collapse=TRUE, fig.height=3,fig.width=4}
# Plot for the 7th area:
plot(results.mod2,unit=7,legend=TRUE)
```

## Neighbors summary


```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=FALSE}
plot(results.mod2, type="neweights")
```


## Lower Saxony Measles Data
 
We now extend the model to add area-specific intercepts (fixed effects) to the endemic component.
 
\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
f.end2 <- addSeason2formula(f = ~ -1 + 
  fe(1, which=rep(TRUE, ncol(measlesWeserEms))) + t, S=1, period=52)
measles.mod3 <- list(ar = list(f = ~ 1), 
     ne = list(f = ~ 1, weights = wji),  
        end = list(f = f.end2, offset = population(measlesWeserEms)),  
            family = "NegBin1")             
results.mod3 <- hhh4(measlesWeserEms, control=measles.mod3)
```

## Lower Saxony Measles Data
 
\scriptsize
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50)}
cbind(results.mod3$coeff,results.mod3$se)
```



## Lower Saxony Measles Data

\tiny
```{r,collapse=TRUE, fig.height=3.0,fig.width=4.5}
plot(results.mod3, unit=7,legend=TRUE)
```

## Residuals

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
y <- measlesWeserEms@observed[2:104,] # Notice no first
time <- matrix(rep(measlesWeserEms@epoch,17),nrow=104,ncol=17)
time <- time[-1,]
#
mu3 <- fitted(results.mod3)
res3 <- (y-mu3)/sqrt(mu3*(1+1.74*mu3))
par(mfrow=c(1,3))
plot(mu3~y,xlab="Observed",ylab="Fitted",type="n")
for (i in 1:17) points(mu3[,i]~y[,i],col=i)
abline(0,1,col="green")
plot(res3~mu3,xlab="Fitted",ylab="Residuals",type="n")
for (i in 1:17) points(res3[,i]~mu3[,i],col=i)
abline(0,0,col="green")
plot(res3~time,xlab="Time (week)",ylab="Residuals",type="n")
for (i in 1:17) points(res3[,i]~time[,i],col=i)
abline(0,0,col="green")
```  

## Residuals

\tiny
```{r,collapse=TRUE, fig.height=3.0,fig.width=4.5,echo=FALSE}
plot(mu3~y,xlab="Observed",ylab="Fitted",type="n")
for (i in 1:17) points(mu3[,i]~y[,i],col=i)
abline(0,1,col="green")
```  

## Residuals

\tiny
```{r,collapse=TRUE, fig.height=3.0,fig.width=4.5,echo=FALSE}
plot(res3~mu3,xlab="Fitted",ylab="Residuals",type="n")
for (i in 1:17) points(res3[,i]~mu3[,i],col=i)
abline(0,0,col="green")

```  


## Residuals

\tiny
```{r,collapse=TRUE, fig.height=3.0,fig.width=4.5,echo=FALSE}
plot(res3~time,xlab="Time (week)",ylab="Residuals",type="n")
for (i in 1:17) points(res3[,i]~time[,i],col=i)
abline(0,0,col="green")
```  


## Random effects model, no neighborhood

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
measles.mod4 <- list(
    ar = list(f = ~ 1),
    end = list(f = addSeason2formula(~0 + ri(type="iid"), S=1, period=52),offset = population(measlesWeserEms)),
    family = "NegBin1")
measles.res4 <- hhh4(measlesWeserEms,measles.mod4)
```
## fitted values for a single unit

```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=FALSE}
plot(measles.res4, units=7)
```

## plot the multiplicative effect of seasonality

```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=FALSE}
plot(measles.res4, type="season")
```

## random intercepts of the endemic component
```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res4, type="ri", component="end", labels=list(cex=0.3, font=3))
```

## Neighbors summary


```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=FALSE}
plot(measles.res4, type="neweights")
```


## Random effects model

Without normalization, power-law weights are $w_{ji} = o_{ji}^{-d}$, where $o_{ji}$ is the order of neighbourhood between regions $i$ and $j$.

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
measles.mod5 <- list(
  end=list(addSeason2formula(~0 + ri(type="iid"), S=1, period=52)), 
  ar=list(f=~1), 
  ne=list(f = ~ -1+ri(type="iid",corr="all"), 
    weights=W_powerlaw(maxlag=5)),family="NegBin1")
measles.res5 <- hhh4(measlesWeserEms, measles.mod5)
```

## Neighborhood weights

Plot the (estimated) neighbourhood weights (``neweights``) as a function of neighbourhood order (shortest-path distance between regions), i.e., $w_{ji} \sim o_{ji}$. 

```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res5, type="neweights")
```

## random intercepts of the endemic component
```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res5,type="fitted",units=c(7,12))
```


## Random effects in all

\small
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
f.end <- addSeason2formula(~0 + ri(type="iid",corr="all"), S=1, period=52)
measles.mod6 <- list(
  end=list(f.end,offset = population(measlesWeserEms)),
  ar=list(f=~-1+ri(type="iid",corr="all")), 
  ne=list(f = ~ -1+ri(type="iid",corr="all"), 
    weights=W_powerlaw(maxlag=5)),family="NegBin1")
measles.res6 <- hhh4(measlesWeserEms, measles.mod6)
```

## Results

\scriptsize
```{r, message=FALSE, collapse=TRUE, tidy=TRUE,tidy.opts=list(width.cutoff=50),fig.show="hide"}
summary(measles.res6)
```


## random intercepts of the endemic component

```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res6,type="fitted",units=c(12))
```


## Random intercepts of the endemic component
```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res6, type="ri", component="ar")
```

## Random intercepts of the endemic component
```{r,collapse=TRUE, fig.height=3,fig.width=4,echo=TRUE}
plot(measles.res6, type="ri", component="ne")
```


