---
title: "dengue"
author: "daniel"
date: "July 18, 2016"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
---

# Problem

Dengue fever is a virus-caused disease that is spread by mosquitoes that a↵ects millions of people in tropical
environments around the Globe. In this problem, you are asked to construct a simple version of the digital
disease detection tool: “Google Dengue Trends” for Mexico. For this, you will download the spreadsheet
Dengue trends AM 111.xls from the course website. The first column in the spreadsheet represents the
date (in months, from 2004-2011), the second column represents the number of Google searches of the term
“dengue” in Mexico, in a given month. The third column represents the number of cases of Dengue in
Mexico, as reported by the Mexican Ministry of Health. You may use Matlab or Excel for this problem.

```{r, warning=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)

# help with knitr and working directories
if (!interactive()) {
    data <- 'Dengue_trends_AM_111.xls'
} else {
    data <- 'sismid_10-statistics_and_modeling_with_novel_data_streams/dengue/Dengue_trends_AM_111.xls'
}

dengue <- read_excel(data)

# weird excel NA values
dengue <- dengue[complete.cases(dengue), ]

# there are spaces in the column names
names(dengue) <- c('date', 'cases', 'searches')
```

```{r}
# code from r graphics cookbook
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# Part A

Plot the number of cases of Dengue as a function of time.

```{r}
head(dengue)
tail(dengue)
```

```{r}
ggplot(data = dengue) + geom_line(aes(x = date, y = cases))
```

```{r}
mod <- glm(formula = cases ~ searches, data = dengue)
summary(mod)
```

```{r}
dengue$glm_predict_all <- predict(mod, dengue[, 'searches'])
ggplot(data = dengue) +
    geom_point(aes(x = searches, y = cases)) +
    geom_line(aes(x = searches, y = glm_predict_all)) +
    ggtitle('Searches by Cases with linear predictor')
```

# Part B

For the training period 2004-2006 (36 months), find the best line that explains the number of cases
of Dengue as a function of the number of searches of the term “dengue”. You should do this by solving
the least squares problem, and you should obtain the value of the y-intercept and the slope.


```{r}
training_df <- dengue[year(dengue$date) >= 2004 & year(dengue$date) <= 2006, ]
mod_training <- glm(cases ~ searches, data = training_df)
summary(mod_training)
```

# Part C

Use the equation of the line you obtained in (b) and plot the number of cases as a function of the number
of searches of the term “dengue”, predicted by your method during the training period. Compare your
results to the plot in (a) for such time period.

```{r}
training_df$values_train <- predict(mod_training, training_df[, 'searches'])

g1 <- ggplot(data = training_df) +
    geom_point(aes(x = searches, y = cases)) +
    geom_line(aes(x = searches, y = values_train)) +
    ggtitle('Searches x Cases 2004 - 2006 (Training)')
g1
```

# Part D

For the prediction or validation period 2007-2011, use the equation of the line you obtained in (b)
to predict the number of the dengue cases as a function of the number of searches of the term “dengue”
from 2007-2011. Plot your predictions and compare them to the actual number of cases.

```{r}
testing_df <- dengue[year(dengue$date) >= 2007, ]
```

```{r}
testing_df$predicted <- predict(mod_training, testing_df[, 'searches'])

g2 <- ggplot(data = testing_df) +
    geom_point(aes(x = searches, y = cases)) +
    geom_line(aes(x = searches, y = predicted)) +
    ggtitle('Searches x Cases 2007 - 2011 (Testing)')
g2
```

```{r}
multiplot(g1 + xlim(-1, 6) + ylim(0, 30000),
          g2 + xlim(-1, 6) + ylim(0, 30000))
```

```{r}
dengue$training_predict <- predict(mod_training, dengue[, 'searches'])
ggplot(data = dengue) +
    geom_line(aes(x = date, y = cases)) +
    geom_line(aes(x = date, y = training_predict), color = 'red')
```

# Part E

Discuss your results. Could you improve this modeling approach? If so, how?

- retrain as new data comes in
- account for auto correlation


