---
title: "mcmc_bayes"
author: "daniel"
date: "July 21, 2016"
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

```{r}
rm(list = ls())

if (interactive()) {
    coding <- 'sismid_11-mcmc2_for_infectious_diseases/lab-02_mcmc-bayes/coding.R'
    mcmc <- 'sismid_11-mcmc2_for_infectious_diseases/lab-02_mcmc-bayes/mcmc-Markov.R'
    data_file <- 'sismid_11-mcmc2_for_infectious_diseases/lab-02_mcmc-bayes/data.txt'
    gamma <- 'sismid_11-mcmc2_for_infectious_diseases/lab-02_mcmc-bayes/mcmc-gamma.R'
} else {
    coding <- 'coding.R'
    mcmc <- 'mcmc-Markov.R'
    data_file <- 'data.txt'
    gamma <- 'mcmc-gamma.R'
}

source(coding)
source(mcmc)
source(gamma)
```



```{r, warning=FALSE, message=FALSE}

```

# Excercise 1

Have a look at the function mcmc-Markov.R and make sure you understand what is going on.

# Excercise 2

Read in the data

```{r}
data <-  read.table(data_file, header = TRUE)
```

# Excercise 3

fit a non-Markovian model to the observed data where the infectious period is assumed to be Gamma distributed with parameters

# Excercise 4

Write a function in R which will draw samples from the posterior distribution of interest, π(β,γ,I|R)π(β,γ,I|R), using MCMC and by making use of the function in the file coding.R.

```{r}
mcmcSIR.Markov(data = data, iter = 10)
```

```{r}
mcmcSIR.gamma(data, 10)
```
