---
title: "lab-01_simulation"
author: "daniel"
date: "July 20, 2016"
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

```{r}
if (interactive()) {
    f <- 'sismid_11-mcmc2_for_infectious_diseases/lab-01_simulation/simulation.R'
    f2 <- 'sismid_11-mcmc2_for_infectious_diseases/lab-01_simulation/simulation_2.R'
} else {
    f <- 'simulation.R'
    f2 <- 'simulation_2.R'
}
```


```{r}
library(ggplot2)
library(tidyr)
library(tibble)

source(f)
source(f2)

set.seed(42)
```


# Exercise 1
Go through the lines of each function and make sure that you follow the logic behind.

# Exercise 2
Simulate realisations from a Markovian SIR using the function `simSIR.Markov` and make sure that you understand the output. You may assume that size of the population size is $N=21$ (i.e. 20 susceptibles and 1 initial infective). In addition, you could try different values for $(β,γ)$, e.g. $(0.9,1)$,$(2,1)$and $(4,1)$.

```{r}
simSIR.Markov(N = 21, beta = 0.9, gamma = 1)
```

```{r}
simSIR.Markov(N = 21, beta = 2, gamma = 1)
```

```{r}
simSIR.Markov(N = 21, beta = 4, gamma = 1)
```

# Exercise 3
Modify the existing functions in simulation.R to record the final size and the duration of the epidemic as part of the functions’ output.

```{r}
simSIR.Markov2(N = 21, beta = 4, gamma = 1)
```


# Exercise 4
Derive a simulation-based estimate of the distribution of the final size of a Markovian SIR model for different values of R0R0, e.g. R0=0.9R0=0.9, R0=1.5R0=1.5 and R0=4R0=4. Furthermore, do the same for the non-Markovian models, e.g. for a constant and a Gamma infectious period. Hint: You may find it useful to write a loop which will iterate the following steps for a number of times:

Simulate a realisation from the epidemic model;
Store the final size
At the end you should have a collection of final sizes for which then you can plot a histogram as your estimate of the true distribution of the final size.

```{r}
sim_results <- list()
for (i in 1:1000) {
    sim <- simSIR.Markov2(N = 100, beta = 4, gamma = 1)
    sim_results <- append(sim_results, list(sim))
}
```

```{r}
num_infected <- sapply(X = sim_results, FUN = '[[', 'num_infected')
# num_infected
```

```{r}
ggplot(data = data.frame(num_infected), aes(x = num_infected)) + geom_histogram() +
    ggtitle('Distribution of number of infected')
```

# Exercise 5
Repeat the above exercise but derive, by simulation, the distribution of the duration of the epidemic instead of the final size.

```{r}
num_events <- sapply(X = sim_results, FUN = '[[', 'num_events')
# num_events
```

```{r}
ggplot(data = data.frame(num_events), aes(x = num_events)) + geom_histogram() +
    ggtitle('distribution of number of events')
```

# Exercise 6
Write a function in R to simulate from a non-Markovian stochastic epidemic model where the infectious period is assumed to follow a Weibull distribution. Hint: The probability density function (pdf) of the Weibull distribution is as follows:

$f(x)=(a/b)(x/b)(a−1)exp(−(x/b)a),x>0,a>0,b>0$

$f(x)=(a/b)(x/b)(a−1)expa(−(x/b)a),x>0,a>0,b>0$

Type `?Weibull` to find out how to simulate from a Weibull distribution.


## Compare markov, gamma, weibull

```{r}
sim_results_markov <- list()
sim_results_gamma <- list()
sim_results_weibull <- list()

sim_N <- 10
sim_b <- 4
sim_g <- 1
sim_d <- 1
total_time <- 1000

for (i in 1:total_time) {
    sim_markov <- simSIR.Markov2(N = sim_N, beta = sim_b, gamma = sim_g)
    sim_results_markov <- append(sim_results_markov, list(sim_markov))
    
    sim_gamma <- simSIR.Non.Markov.gamma2(N = sim_N, beta = sim_b, gamma = sim_g, delta = sim_d)
    sim_results_gamma <- append(sim_results_gamma, list(sim_gamma))
    
    sim_weibull <- simSIR.Non.Markov.weibull(N = sim_N, beta = sim_b, gamma = sim_g, delta = sim_d,
                                             shape = 1, scale = 1)
    sim_results_weibull <- append(sim_results_weibull, list(sim_weibull))
}
```

### Num infected

```{r}
num_infected_m <- sapply(X = sim_results_markov, FUN = '[[', 'num_infected')
num_infected_g <- sapply(X = sim_results_gamma, FUN = '[[', 'num_infected')
num_infected_w <- sapply(X = sim_results_weibull, FUN = '[[', 'num_infected')
```

```{r}

num_infected_df <- data.frame(num_infected_m, num_infected_g, num_infected_w)
num_infected_df <- num_infected_df %>%
    gather(key = dist, value = value)

num_infected_df$time <- rep(1:total_time, times = 3)
```


```{r}
ggplot(data = num_infected_df, aes(x = value)) + geom_histogram() +
    facet_wrap(~dist) + 
    ggtitle('Distribution of number of infected')
```

### Num events

```{r}
num_events_m <- sapply(X = sim_results_markov, FUN = '[[', 'num_events')
num_events_g <- sapply(X = sim_results_gamma, FUN = '[[', 'num_events')
num_events_w <- sapply(X = sim_results_weibull, FUN = '[[', 'num_events')
```

```{r}

num_events_df <- data.frame(num_events_m, num_events_g, num_events_w)
num_events_df <- num_events_df %>%
    gather(key = dist, value = value)

num_events_df$time <- rep(1:total_time, times = 3)
```


```{r}
ggplot(data = num_events_df, aes(x = value)) + geom_histogram() +
    facet_wrap(~dist) + 
    ggtitle('Distribution of number of events')
```

# Exercise 7
Write a function to simulate from an epidemic model which involves a fixed latent period, i.e. write a function to simulate from a stochastic SEIR model.

```{r}
simSIR.Non.Markov.constant(N = 25, beta = 32, k = 10)
```

